```{r setup}
library(readxl)
library(data.table)
#-----------------------------------------------------------------------
setwd("/Users/dinhngockhanh/CancerSimulator/R/")
files_sources <- list.files(pattern = "*.r$")
sapply(files_sources, source)
#-----------------------------------------------------------------------
# devtools::install_github("dinhngockhanh/CancerSimulator")
# library(CancerSimulator)
```

#   Create simulated arm-level library with CancerSimulator

First we set up the fixed parameters:

```{r}
cell_lifespan <- 30
T_0 <- list(0, "year")
T_end <- list(80, "year")
Table_sample <- data.frame(Sample_ID = c("SA01"), Cell_count = c(Inf), Age_sample = c(80))
T_tau_step <- cell_lifespan / 2
CN_bin_length <- 500000
selection_model <- "chrom-arm-selection"
#------------------------------------------------------CNA PROBABILITIES
prob_CN_whole_genome_duplication <- 0e-4
prob_CN_missegregation <- 5e-5
prob_CN_chrom_arm_missegregation <- 0e-5
prob_CN_focal_amplification <- 0e-5
prob_CN_focal_deletion <- 0e-5
prob_CN_cnloh_interstitial <- 0e-5
prob_CN_cnloh_terminal <- 0e-5
model_CN_focal_amplification_length <- "beta"
model_CN_focal_deletion_length <- "beta"
prob_CN_focal_amplification_length_shape_1 <- 0.758304780825031
prob_CN_focal_amplification_length_shape_2 <- 5.33873409782625
prob_CN_focal_deletion_length_shape_1 <- 0.814054548726361
prob_CN_focal_deletion_length_shape_2 <- 6.16614890284825
prob_CN_cnloh_interstitial_length <- 0.005
prob_CN_cnloh_terminal_length <- 0.005
rate_driver <- 0
rate_passenger <- 1e-11
#---------------------------------------------------VIABILITY THRESHOLDS
bound_driver <- Inf
bound_average_ploidy <- Inf
bound_homozygosity <- 0
bound_maximum_CN <- Inf
bound_maximum_CN_normalized <- 4
#----------------------------------------------------POPULATION DYNAMICS
vec_time <- T_0[[1]]:T_end[[1]]
L <- 10000
t_0 <- 20
k <- 0.3
vec_cell_count <- L / (1 + exp(-k * (vec_time - t_0)))
table_population_dynamics <- cbind(vec_time, vec_cell_count)
#-----------------------------------------------------------------------
gc <- read.csv(file = system.file("extdata", "gc_map_500kb.csv", package = "CancerSimulator"))
gc_slope <- 1.2
gc_int <- 0
sigma1 <- 0.1
num_reads <- 2e6
model_variables <- BUILD_general_variables(
    cell_lifespan = cell_lifespan,
    T_0 = T_0, T_end = T_end, T_tau_step = T_tau_step,
    Table_sample = Table_sample,
    CN_bin_length = CN_bin_length,
    prob_CN_whole_genome_duplication = prob_CN_whole_genome_duplication,
    prob_CN_missegregation = prob_CN_missegregation,
    prob_CN_chrom_arm_missegregation = prob_CN_chrom_arm_missegregation,
    prob_CN_focal_amplification = prob_CN_focal_amplification,
    prob_CN_focal_deletion = prob_CN_focal_deletion,
    prob_CN_cnloh_interstitial = prob_CN_cnloh_interstitial,
    prob_CN_cnloh_terminal = prob_CN_cnloh_terminal,
    model_CN_focal_amplification_length = model_CN_focal_amplification_length,
    model_CN_focal_deletion_length = model_CN_focal_deletion_length,
    prob_CN_focal_amplification_length_shape_1 = prob_CN_focal_amplification_length_shape_1,
    prob_CN_focal_amplification_length_shape_2 = prob_CN_focal_amplification_length_shape_2,
    prob_CN_focal_deletion_length_shape_1 = prob_CN_focal_deletion_length_shape_1,
    prob_CN_focal_deletion_length_shape_2 = prob_CN_focal_deletion_length_shape_2,
    prob_CN_cnloh_interstitial_length = prob_CN_cnloh_interstitial_length,
    prob_CN_cnloh_terminal_length = prob_CN_cnloh_terminal_length,
    rate_driver = rate_driver,
    rate_passenger = rate_passenger,
    selection_model = selection_model,
    bound_driver = bound_driver,
    bound_maximum_CN = bound_maximum_CN,
    bound_average_ploidy = bound_average_ploidy,
    bound_homozygosity = bound_homozygosity,
    bound_maximum_CN_normalized = bound_maximum_CN_normalized,
    table_population_dynamics = table_population_dynamics,
    gc = gc,
    gc_slope = gc_slope,
    gc_int = gc_int,
    sigma1 = sigma1,
    num_reads = num_reads
)
arm_id <- c(paste(model_variables$cn_info$Chromosome, "p", sep = ""), paste(model_variables$cn_info$Chromosome, "q", sep = ""))
arm_chromosome <- rep(model_variables$cn_info$Chromosome, 2)
arm_start <- c(rep(1, length(model_variables$cn_info$Chromosome)), model_variables$cn_info$Centromere_location + 1)
arm_end <- c(model_variables$cn_info$Centromere_location, model_variables$cn_info$Bin_count)
arm_s <- rep(1, length(arm_id))

model_variables <- BUILD_driver_library(
    model_variables = model_variables,
    table_arm_selection_rates = data.frame(Arm_ID = arm_id, Chromosome = arm_chromosome, Bin_start = arm_start, Bin_end = arm_end, s_rate = arm_s)
)

cell_count <- 20
CN_matrix <- BUILD_cn_normal_autosomes(model_variables$cn_info)
drivers <- list()
model_variables <- BUILD_initial_population(model_variables = model_variables, cell_count = cell_count, CN_matrix = CN_matrix, drivers = drivers)
library_name <- "ABC-BULK-ARM-CN"
model_variables <- CHECK_model_variables(model_variables)
```

We create a list of all parameters to be fitted, their types (CNA probabilities or chromosome-arm selection rates), and the bounds for their respective uniform prior distributions:

```{r}
list_parameters_library <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(list_parameters_library) <- c("Variable", "Type", "Lower_bound", "Upper_bound")
list_parameters_library[nrow(list_parameters_library) + 1, ] <- c("prob_CN_chrom_arm_missegregation", "CNA_probability", 1e-5, 1e-4)
for (i in 1:length(model_variables$chromosome_arm_library$Arm_ID)) {
    list_parameters_library[nrow(list_parameters_library) + 1, ] <- c(
        model_variables$chromosome_arm_library$Arm_ID[i],
        "Arm_selection_rate",
        0.5, 1.5
    )
}
```

We create a list for the statistics targets. This would be the list of chromosome arms. The CN gain/loss frequencies on each arm form the statistics for ABC.

```{r}
list_targets_library <- model_variables$chromosome_arm_library$Arm_ID
```

We then create a simulated arm-level library with CancerSimulator:

```{r}
# start_time <- Sys.time()
# library_bulk_arm_CN(
#     library_name = library_name,
#     model_variables = model_variables,
#     list_parameters = list_parameters_library,
#     list_targets = list_targets_library,
#     ABC_simcount = 1000,
#     n_samples = 100
# )
# end_time <- Sys.time()
# print(end_time - start_time)
```

#   Fit the Copy Number data for pan-cancer from TCGA

We first load the pan-cancer arm-level gain/loss data from TCGA:

```{r}
# copynumber_DATA <- read.csv(file = system.file("extdata", "Davoli_Charm_score.csv", package = "CancerSimulator"))
# copynumber_DATA <- copynumber_DATA[, c(1, 2, 3, 14, 15)]
# colnames(copynumber_DATA) <- c("Arm", "Del_freq_all", "Amp_freq_all", "Charm.TSG.OG.score", "Charm.TSG.OG.Ess.score")
```

Limit the list of parameters to chromosome arms with available data:

```{r}
# list_targets <- copynumber_DATA$Arm
# list_parameters <- list_parameters_library[c(
#     which(list_parameters_library$Type == "CNA_probability"),
#     which(list_parameters_library$Type == "Arm_selection_rate" & list_parameters_library$Variable %in% copynumber_DATA$Arm)
# ), ]
```

We then run the fitting routine:

```{r}
# cat(paste0("\n\n\nFITTING FOR PAN-CANCER\n"))
# fitting_bulk_arm_CN(
#     library_name = library_name,
#     model_name = "PAN-CANCER",
#     model_variables = model_variables,
#     copynumber_DATA = copynumber_DATA,
#     list_parameters = list_parameters,
#     list_parameters_library = list_parameters_library,
#     list_targets = list_targets,
#     list_targets_library = list_targets_library,
#     N_shuffle = 19,
#     type_sample_DATA = "average",
#     type_cn_DATA = "arm",
#     folder_workplace = library_name
# )
```

#   Fit the Copy Number data for each cancer type from PCAWG

We first load the list of samples for all cancer types in PCAWG:

```{r}
copynumber_PCAWG <- read_excel(system.file("pcawg_sample_sheet.xlsx", package = "CancerSimulator"))
```

List of cancer type-specific datasets in PCAWG:

```{r}
PCAWG_cancer_types <- unique(copynumber_PCAWG$dcc_project_code)
```

WGD status for each sample in PCAWG:

```{r}
PCAWG_wgd <- read.table(system.file("consensus.20170218.purity.ploidy.txt", package = "CancerSimulator"), header = TRUE)
```

List of all samples with available high-quality CN data, per PCAWG cancer type:

```{r}
PCAWG_cancer_type_sample_ids <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    sample_ids <- copynumber_PCAWG$aliquot_id[which(copynumber_PCAWG$dcc_project_code == PCAWG_cancer_types[i])]
    for (j in 1:length(sample_ids)) {
        sample_id <- sample_ids[j]
        if (!file.exists(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"))) next
        tmp <- read.table(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"), header = TRUE)
        if (max(tmp$star, na.rm = TRUE) < 2) next
        PCAWG_cancer_type_sample_ids[[i]] <- c(PCAWG_cancer_type_sample_ids[[i]], sample_id)
    }
}
```

List of only non-WGD samples with available high-quality CN data, per PCAWG cancer type:

```{r}
PCAWG_cancer_type_nonWGD_sample_ids <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    sample_ids <- PCAWG_cancer_type_sample_ids[[i]]
    sample_wgd <- PCAWG_wgd[which(PCAWG_wgd$samplename %in% sample_ids),]
    sample_nonwgd_ids <- sample_wgd$samplename[which(sample_wgd$wgd_uncertain==FALSE & sample_wgd$wgd_status=='no_wgd')]
    PCAWG_cancer_type_nonWGD_sample_ids[[i]]<-sample_nonwgd_ids
}
```

Remove PCAWG cancer types with too few samples:

```{r}
# sample_sizes <- sapply(PCAWG_cancer_type_sample_ids, length)
# PCAWG_cancer_types <- PCAWG_cancer_types[-which(sample_sizes < 30)]
# PCAWG_cancer_type_sample_ids <- PCAWG_cancer_type_sample_ids[-which(sample_sizes < 30)]
```

Remove PCAWG cancer types with less than 30 non-WGD samples:

```{r}
sample_sizes <- sapply(PCAWG_cancer_type_nonWGD_sample_ids, length)
PCAWG_cancer_types <- PCAWG_cancer_types[-which(sample_sizes < 30)]
PCAWG_cancer_type_sample_ids <- PCAWG_cancer_type_sample_ids[-which(sample_sizes < 30)]
PCAWG_cancer_type_nonWGD_sample_ids <- PCAWG_cancer_type_nonWGD_sample_ids[-which(sample_sizes < 30)]
```

Combine CN profiles for all samples of each PCAWG cancer type into one:

```{r}
copynumber_DATA_cancer_types_nonWGD <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    cancer_type <- PCAWG_cancer_types[i]
    cancer_type_sample_ids <- PCAWG_cancer_type_nonWGD_sample_ids[[i]]
    copynumber_DATA_ls <- vector("list", length(cancer_type_sample_ids))
    for (j in 1:length(cancer_type_sample_ids)) {
        sample_id <- cancer_type_sample_ids[j]
        tmp <- read.table(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"), header = TRUE)
        if (any(is.na(tmp$star))) tmp <- tmp[-which(is.na(tmp))]
        tmp <- tmp[which(tmp$star >= 2), ]
        tmp$donor_unique_id <- sample_id
        copynumber_DATA_ls[[j]] <- tmp
    }
    copynumber_DATA_cancer_types_nonWGD[[i]] <- rbindlist(copynumber_DATA_ls)
}

copynumber_DATA_cancer_types <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    cancer_type <- PCAWG_cancer_types[i]
    cancer_type_sample_ids <- PCAWG_cancer_type_sample_ids[[i]]
    copynumber_DATA_ls <- vector("list", length(cancer_type_sample_ids))
    for (j in 1:length(cancer_type_sample_ids)) {
        sample_id <- cancer_type_sample_ids[j]
        tmp <- read.table(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"), header = TRUE)
        if (any(is.na(tmp$star))) tmp <- tmp[-which(is.na(tmp))]
        tmp <- tmp[which(tmp$star >= 2), ]
        tmp$donor_unique_id <- sample_id
        copynumber_DATA_ls[[j]] <- tmp
    }
    copynumber_DATA_cancer_types[[i]] <- rbindlist(copynumber_DATA_ls)
}
```

```{r}
list_targets <- model_variables$chromosome_arm_library$Arm_ID
list_parameters <- list_parameters_library
```

Fit the chromosome-arm-level gain/loss frequencies with ABC random forest:

```{r}
for (i in 1:length(PCAWG_cancer_types)) {
    cancer_type <- PCAWG_cancer_types[i]
    copynumber_DATA <- copynumber_DATA_cancer_types_nonWGD[[i]]
    cat(paste0("\n\n\nFITTING FOR ", cancer_type, "   [", i, "/", length(PCAWG_cancer_types), "]", "\n"))
    fitting_bulk_arm_CN(
        library_name = library_name,
        model_name = cancer_type,
        model_variables = model_variables,
        copynumber_DATA = copynumber_DATA,
        list_parameters = list_parameters,
        list_parameters_library = list_parameters_library,
        list_targets = list_targets,
        list_targets_library = list_targets_library,
        bound_freq = 0.1,
        library_shuffle = TRUE,
        folder_workplace = library_name
    )
}
```

#   Statistics for PCAWG cancer types

```{r}
df_WGD_FGA <- statistics_bulk_arm_WGD_status(
    plotname = 'ALL_PCAWG',
    DATA_cancer_types = PCAWG_cancer_types,
    DATA_cancer_type_sample_ids = PCAWG_cancer_type_sample_ids,
    DATA_cancer_type_cn = copynumber_DATA_cancer_types,
    DATA_wgd = PCAWG_wgd,
    model_variables = model_variables
)
```

#   Fit the WGD proportion and WGD+ CNA rates for each cancer type from PCAWG

Update the formulae for probabilities of CNAs per cell division. The formulae are linear with respect to DNA length (as proxy for level of aneuploidy). The slope of the relationship (i.e. how much CNA rates change as a result of distinct ploidy) is subject for fitting.

```{r}
model_variables$general_variables[nrow(model_variables$general_variables) + 1, ] <- c("formula_CN_slope", 1, "", "")



DNA_length_original <- 2 * sum(model_variables$cn_info$Bin_count) *
    as.numeric(model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "size_CN_block_DNA")])
model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_missegregation")] <-
    paste0("prob_CN_missegregation * (DNA_length / ", DNA_length_original, ")^formula_CN_slope")
model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_chrom_arm_missegregation")] <-
    paste0("prob_CN_chrom_arm_missegregation * (DNA_length / ", DNA_length_original, ")^formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_missegregation")] <-
#     paste0("prob_CN_missegregation * DNA_length / ", DNA_length_original, " * formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_chrom_arm_missegregation")] <-
#     paste0("prob_CN_chrom_arm_missegregation * DNA_length / ", DNA_length_original, " * formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_focal_amplification")] <-
#     paste0("prob_CN_focal_amplification * DNA_length / ", DNA_length_original, " * formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_focal_deletion")] <-
#     paste0("prob_CN_focal_deletion * DNA_length / ", DNA_length_original, " * formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_cnloh_interstitial")] <-
#     paste0("prob_CN_cnloh_interstitial * DNA_length / ", DNA_length_original, " * formula_CN_slope")
# model_variables$general_variables$Value[which(model_variables$general_variables$Variable == "formula_CN_cnloh_terminal")] <-
#     paste0("prob_CN_cnloh_terminal * DNA_length / ", DNA_length_original, " * formula_CN_slope")
```

The list of parameters to fit for each cancer type includes the slope of relationship between CNA rates and cell ploidy, and probability of WGD per cell division. Similarly to previous fitting, we use uniform prior distributions:

```{r}
list_parameters_cancer_type_library <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(list_parameters_cancer_type_library) <- c("Variable", "Type", "Lower_bound", "Upper_bound")
list_parameters_cancer_type_library[nrow(list_parameters_cancer_type_library) + 1, ] <- c("prob_CN_whole_genome_duplication", "CNA_probability", 1e-5, 1e-4)
list_parameters_cancer_type_library[nrow(list_parameters_cancer_type_library) + 1, ] <- c("formula_CN_slope", "Extra_parameter", 0, 3)
```








For each cancer type, the fitted CNA probabilities and chromosome-arm selection rates are first assigned into the model variables. Then a library of simulations is created, where target statistics are WGD percentage, arm-level CN deviation from 2 in WGD-negative cells, and arm-level CN deviation from 4 in WGD-positive cells. Finally, the cancer type-specific library is used to fit the parameters.

```{r}
for (i in 1:length(PCAWG_cancer_types)) {
    cancer_type <- PCAWG_cancer_types[i]
    copynumber_DATA <- copynumber_DATA_cancer_types[[i]]
    samples_DATA <- PCAWG_cancer_type_sample_ids[[i]]
    wgd_uncertain_DATA <- PCAWG_wgd$wgd_uncertain[which(PCAWG_wgd$samplename %in% samples_DATA)]
    wgd_status_DATA <- PCAWG_wgd$wgd_status[which(PCAWG_wgd$samplename %in% samples_DATA)]
    if (length(which(wgd_uncertain_DATA == FALSE & wgd_status_DATA == "no_wgd")) == 0) {
        next
    } else if (length(which(wgd_uncertain_DATA == FALSE & wgd_status_DATA == "wgd")) == 0) {
        next
    }
    library_name_cancer_type <- paste0("ABC-BULK-ARM-WGD-", cancer_type)



    if (cancer_type!='PACA-AU') next



    parameter_best <- read.csv(paste0(cancer_type, "_fitted_parameters.csv"))
    model_variables_cancer_type <- bulk_arm_CN_assign_paras(
        model_variables = model_variables,
        parameter_IDs = parameter_best$Variable,
        parameters = parameter_best$Best_value
    )



    # start_time <- Sys.time()
    # library_bulk_arm_WGD(
    #     library_name = library_name_cancer_type,
    #     model_variables = model_variables_cancer_type,
    #     list_parameters = list_parameters_cancer_type_library,
    #     list_targets = list_targets,
    #     ABC_simcount = 1000,
    #     n_samples = 100
    # )
    # end_time <- Sys.time()
    # print(end_time - start_time)



    # fitting_bulk_arm_WGD(
    #     library_name = library_name_cancer_type,
    #     model_name = paste0(cancer_type,'_WGD'),
    #     model_variables = model_variables,
    #     copynumber_DATA = copynumber_DATA,
    #     wgd_DATA = PCAWG_wgd,
    #     list_parameters = list_parameters_cancer_type_library,
    #     list_parameters_library = list_parameters_cancer_type_library,
    #     list_targets = list_targets,
    #     # list_targets_library = list_targets_library,
    #     folder_workplace = library_name_cancer_type
    # )



    model_variables_cancer_type$general_variables$Value[which(model_variables_cancer_type$general_variables$Variable=='prob_CN_whole_genome_duplication')] <- 1e-6
    model_variables_cancer_type$general_variables$Value[which(model_variables_cancer_type$general_variables$Variable=='formula_CN_slope')] <- log2(100)
    tmp <- simulator_full_program(
        model = model_variables_cancer_type,
        n_simulations = 100,
        stage_final = 1,
        save_simulation = FALSE,
        report_progress = TRUE,
        # seed = 11,
        compute_parallel = FALSE
    )
}
```












```{r}
```
