---
title: "Chromosome Arm Fitting Program"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chromosome Arm Fitting Program}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    warning = FALSE, message = FALSE,
    comment = "#>"
)
```


Five vectors are used to fit the chromosome arm selection rates. Given values at the same location in the vectors, `arm_s` is the selection rate for the arm with ID `arm_id`, which runs from starting bin `arm_start` to ending bin `arm_end` on chromosome `arm_chromosome`. The selection rate is initialized to be 1 for each chromosome arm in `arm_ID`.

```{r}
arm_id <- c(paste(model_variables$cn_info$Chromosome, "p", sep = ""), paste(model_variables$cn_info$Chromosome, "q", sep = ""))
arm_chromosome <- rep(model_variables$cn_info$Chromosome, 2)
arm_start <- c(rep(1, length(model_variables$cn_info$Chromosome)), model_variables$cn_info$Centromere_location + 1)
arm_end <- c(model_variables$cn_info$Centromere_location, model_variables$cn_info$Bin_count)
arm_s <- rep(1, length(arm_id))
```

After initializing the selection rate for each chromosome arm, we can update `model_variables` with the variables for the selection model. Include only the tables required for the selected selection model:

```{r}
model_variables <- BUILD_driver_library(
    model_variables = model_variables,
    table_arm_selection_rates = data.frame(Arm_ID = arm_id, Chromosome = arm_chromosome, Bin_start = arm_start, Bin_end = arm_end, s_rate = arm_s)
)
```

## Initial cell population

We now specify the initial cell population. Here we assume that the initial state consists of one clone, where `cell_count` is its cell count. The variable `CN_matrix` specifies the clone's CN profile, here assumed to be normal diploid female cells. The variable `drivers` defines the existing driver mutations in the clone, here assumed to be none.

```{r}
cell_count <- 20
CN_matrix <- BUILD_cn_normal_autosomes(model_variables$cn_info)
drivers <- list()
model_variables <- BUILD_initial_population(model_variables = model_variables, cell_count = cell_count, CN_matrix = CN_matrix, drivers = drivers)
```

???????????????????????????????????????????????
The final step of model variable preparation is using the function `CHECK_model_variables` to check the variables defined thus far in csv files. The variable `library_name` is incorporated into the csv files' names.

```{r}
library_name <- "ABC-BULK-ARM-CN"
model_variables <- CHECK_model_variables(model_variables)
```
???????????????????????????????????????????????

`list_parameters_library` is a dataframe consisting of the parameters (`variable`) to be fitted: `prob_CN_chrom_arm_missegregation` and the selection rate for each chromosome arm. `prob_CN_chrom_arm_missegregation` has lower and upper bounds of 10^-5^ and 10^-4^, respectively. Each chromosome arm has lower and upper bounds of 0.5 and 1.5, respectively.

```{r}
list_parameters_library <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(list_parameters_library) <- c("Variable", "Type", "Lower_bound", "Upper_bound")
list_parameters_library[nrow(list_parameters_library) + 1, ] <- c("prob_CN_chrom_arm_missegregation", "CNA_probability", 1e-5, 1e-4)
for (i in 1:length(model_variables$chromosome_arm_library$Arm_ID)) {
    list_parameters_library[nrow(list_parameters_library) + 1, ] <- c(
        model_variables$chromosome_arm_library$Arm_ID[i],
        "Arm_selection_rate",
        0.5, 1.5
    )
}
```

Initialize a vector specifying the variables we want to fit.

```{r}
list_targets_library <- model_variables$chromosome_arm_library$Arm_ID
```

Here, we build a library of simulations using our specified model variables. Our variables to fit are specified by the `list_targets` parameter. To generate statistics necessary for ABC, `n_sample` is the number of values from the prior distribution that are randomly selected for each simulation. `ABC_simcount` is then the number of simulations for a given parameters set. Each simulation is characterized by the copy number profile of the larest clone. The gain/loss frequencies of each chromosome arm are the important statistic used later for fitting arm selection rates.

```{r}
###################################    BUILD LIBRARY OF SIMULATED ARM-CN
start_time <- Sys.time()
library_bulk_arm_CN(
    library_name = library_name,
    model_variables = model_variables,
    list_parameters = list_parameters_library,
    list_targets = list_targets_library,
    ABC_simcount = 2000,
    n_samples = 100,
    R_libPaths = "/burg/iicd/users/knd2127/rpackages"
)
end_time <- Sys.time()
print(end_time - start_time)
########################################################################
```

We then aggregate the arm deletion, arm amplification, and gene balance scores betweem tumor supressor, onco, and essential genes from TCGA pan-cancer data.

```{r}
#############################################    FIT FOR TCGA PAN-CANCER
copynumber_DATA <- read.csv(file = system.file("extdata", "Davoli_Charm_score.csv", package = "CancerSimulator"))
copynumber_DATA <- copynumber_DATA[, c(1, 2, 3, 14, 15)]
colnames(copynumber_DATA) <- c("Arm", "Del_freq_all", "Amp_freq_all", "Charm.TSG.OG.score", "Charm.TSG.OG.Ess.score")
```

Because TCGA pan-cancer excludes chromosome arms 13p, 14p, 15p, 21q, 22p, and chromosomes X and Y, we need to modify the list of parameters that need to be fit by changing the value of `list_targets` and `list_parameters`. 

```{r}
list_targets <- copynumber_DATA$Arm
list_parameters <- list_parameters_library[c(
    which(list_parameters_library$Type == "CNA_probability"),
    which(list_parameters_library$Type == "Arm_selection_rate" & list_parameters_library$Variable %in% copynumber_DATA$Arm)
), ]
```

Fitting routine for each chromosome arm using ABCrf (and bootstrapping). ???more detail??? (Supplement 1)

```{r}
cat(paste0("\n\n\nFITTING FOR PAN-CANCER\n"))
fitting_bulk_arm_CN(
    library_name = library_name,
    model_name = "PAN-CANCER",
    model_variables = model_variables,
    copynumber_DATA = copynumber_DATA,
    list_parameters = list_parameters,
    list_parameters_library = list_parameters_library,
    list_targets = list_targets,
    list_targets_library = list_targets_library,
    library_shuffle = TRUE,
    type_sample_DATA = "average",
    type_cn_DATA = "arm",
    folder_workplace = library_name,
    R_libPaths = "/burg/iicd/users/knd2127/rpackages"
)
########################################################################
```

For fitting arm selection rates by cancer subtype in PCAWG data, we retrieve the different types of cancer present in PCAWG data and save it to `PCAWG_cancer_types`.

```{r}
#######################################    FIT FOR PCAWG CANCER SUBTYPES
copynumber_PCAWG <- read_excel(system.file("pcawg_specimen_histology_August2016_v9.xlsx", package = "CancerSimulator"))

PCAWG_cancer_types <- unique(copynumber_PCAWG$histology_abbreviation)
PCAWG_cancer_types <- PCAWG_cancer_types[which(!is.na(PCAWG_cancer_types))]
```



```{r}
PCAWG_wgd <- read.table(system.file("consensus.20170218.purity.ploidy.txt", package = "CancerSimulator"), header = TRUE)
```



```{r}
PCAWG_cancer_type_sample_ids <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    sample_ids <- copynumber_PCAWG$tcga_sample_uuid[which(copynumber_PCAWG$histology_abbreviation == PCAWG_cancer_types[i])]
    for (j in 1:length(sample_ids)) {
        sample_id <- sample_ids[j]
        if (!file.exists(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"))) next
        tmp <- read.table(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"), header = TRUE)
        if (max(tmp$star, na.rm = TRUE) < 2) next
        PCAWG_cancer_type_sample_ids[[i]] <- c(PCAWG_cancer_type_sample_ids[[i]], sample_id)
    }
}
```


PCAWG_cancer_type_nonWGD_sample_ids <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    sample_ids <- PCAWG_cancer_type_sample_ids[[i]]
    sample_wgd <- PCAWG_wgd[which(PCAWG_wgd$samplename %in% sample_ids), ]
    sample_nonwgd_ids <- sample_wgd$samplename[which(sample_wgd$wgd_uncertain == FALSE & sample_wgd$wgd_status == "no_wgd")]
    PCAWG_cancer_type_nonWGD_sample_ids[[i]] <- sample_nonwgd_ids
}

sample_sizes <- sapply(PCAWG_cancer_type_nonWGD_sample_ids, length)
PCAWG_cancer_types <- PCAWG_cancer_types[-which(sample_sizes < 10)]
PCAWG_cancer_type_sample_ids <- PCAWG_cancer_type_sample_ids[-which(sample_sizes < 10)]
PCAWG_cancer_type_nonWGD_sample_ids <- PCAWG_cancer_type_nonWGD_sample_ids[-which(sample_sizes < 10)]

copynumber_DATA_cancer_types <- vector("list", length(PCAWG_cancer_types))
for (i in 1:length(PCAWG_cancer_types)) {
    cancer_type <- PCAWG_cancer_types[i]
    cancer_type_sample_ids <- PCAWG_cancer_type_nonWGD_sample_ids[[i]]
    copynumber_DATA_ls <- vector("list", length(cancer_type_sample_ids))
    for (j in 1:length(cancer_type_sample_ids)) {
        sample_id <- cancer_type_sample_ids[j]
        tmp <- read.table(system.file(paste0("consensus.20170119.somatic.cna.annotated/", sample_id, ".consensus.20170119.somatic.cna.annotated.txt"), package = "CancerSimulator"), header = TRUE)
        if (any(is.na(tmp$star))) tmp <- tmp[-which(is.na(tmp))]
        tmp <- tmp[which(tmp$star >= 2), ]
        tmp$donor_unique_id <- sample_id
        copynumber_DATA_ls[[j]] <- tmp
    }
    copynumber_DATA_cancer_types[[i]] <- rbindlist(copynumber_DATA_ls)
}

list_targets <- model_variables$chromosome_arm_library$Arm_ID
list_parameters <- list_parameters_library

for (i in 1:6) {
    cancer_type <- PCAWG_cancer_types[i]
    copynumber_DATA <- copynumber_DATA_cancer_types[[i]]
    cat(paste0("\n\n\nFITTING FOR ", cancer_type, "   [", i, "/", length(PCAWG_cancer_types), "]", "\n"))
    fitting_bulk_arm_CN(
        library_name = library_name,
        model_name = cancer_type,
        model_variables = model_variables,
        copynumber_DATA = copynumber_DATA,
        list_parameters = list_parameters,
        list_parameters_library = list_parameters_library,
        list_targets = list_targets,
        list_targets_library = list_targets_library,
        bound_freq = 0.1,
        library_shuffle = TRUE,
        folder_workplace = library_name,
        R_libPaths = "/burg/iicd/users/knd2127/rpackages"
    )
}
########################################################################



################################    STATISTICS FOR PCAWG CANCER SUBTYPES
df_WGD_FGA <- statistics_bulk_arm_WGD_status(
    plotname = "ALL_PCAWG",
    DATA_cancer_types = PCAWG_cancer_types,
    DATA_cancer_type_sample_ids = PCAWG_cancer_type_sample_ids,
    DATA_cancer_type_cn = copynumber_DATA_cancer_types,
    DATA_wgd = PCAWG_wgd,
    model_variables = model_variables
)
########################################################################