---
title: "How to use Cancer Simulator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use Cancer Simulator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, message = FALSE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(ggtree)
library(signals)
library(dendextend)
library(fishplot)
library(ctc)
library(adephylo)
library(data.table)
library(HMMcopy)
library(plyr)
library(matrixStats)
library(parallel)
library(RColorBrewer)
library(ape)

library(abc)
#-----------------------------------------------------------------------
setwd("/Users/dinhngockhanh/CancerSimulator/R/")
files_sources <- list.files(pattern = "*.r$")
sapply(files_sources, source)
#-----------------------------------------------------------------------
# devtools::install_github("dinhngockhanh/CancerSimulator")
# library(CancerSimulator)
```

# Analyze the noise in DLP+ data

Here we analyze the noise level in `ov2295_cell_cn.csv`, from https://zenodo.org/record/3445364#.Ytq9ji-B3Pw.

```{r}
# ov2295_cn       <<- read.csv(file = 'ov2295_cell_cn.csv')
# vec_gc          <<- read.csv(file = system.file("extdata", "gc_map_500kb.csv", package = "CancerSimulator"))
# analysis_dlp_cn(CNbins=ov2295_cn,gc=gc,filename_prefix='OV2295')
```

# Create one simulation

Save parameters to build one simulation:

```{r}
cell_lifespan                       <- 30
T_0                                 <- list(0,'year')
T_end                               <- list(80,'year')
T_tau_step                          <- cell_lifespan/2
CN_bin_length                       <- 500000
########################################################################
########################################################################
########################################################################
Table_sample                        <- data.frame(Sample_ID='SA01',Cell_count=300,Age_sample=80)



prob_CN_whole_genome_duplication    <- 3.5e-5
prob_CN_missegregation              <- 8e-4
prob_CN_chrom_arm_missegregation    <- 8e-4
prob_CN_focal_amplification         <- 8e-4
prob_CN_focal_deletion              <- 0e-4
prob_CN_cnloh_interstitial          <- 0e-4
prob_CN_cnloh_terminal              <- 0e-4

# prob_CN_whole_genome_duplication    <- 1e-4
# prob_CN_missegregation              <- 7e-4
# prob_CN_chrom_arm_missegregation    <- 2e-4
# prob_CN_focal_amplification         <- 7e-4
# prob_CN_focal_deletion              <- 2e-4
# prob_CN_cnloh_interstitial          <- 4e-4
# prob_CN_cnloh_terminal              <- 6e-4



prob_CN_focal_amplification_length  <- 0.1
prob_CN_focal_deletion_length       <- 0.1
prob_CN_cnloh_interstitial_length   <- 0.1
prob_CN_cnloh_terminal_length       <- 0.1

# prob_CN_focal_amplification_length  <- 0.5
# prob_CN_focal_deletion_length       <- 0.5
# prob_CN_cnloh_interstitial_length   <- 0.5
# prob_CN_cnloh_terminal_length       <- 0.5
########################################################################
########################################################################
########################################################################
rate_driver                         <- 8e-16
rate_passenger                      <- 1e-11
selection_model                     <- 'ancient'
bound_driver                        <- 3
bound_average_ploidy                <- 4.5
bound_homozygosity                  <- 500
vec_time                            <- T_0[[1]]:T_end[[1]]
L                                   <- 10000
t_0                                 <- 20
k                                   <- 0.3
vec_cell_count                      <- L/(1+exp(-k*(vec_time-t_0)))
table_population_dynamics           <- cbind(vec_time,vec_cell_count)
gc                                  <- read.csv(file = system.file("extdata", "gc_map_500kb.csv", package = "CancerSimulator"))
gc_slope                            <- 1.2
gc_int                              <- 0
sigma1                              <- 0.1
num_reads                           <- 2e6
model_variables                     <- BUILD_general_variables(
    cell_lifespan=cell_lifespan,
    T_0=T_0,T_end=T_end,T_tau_step=T_tau_step,
    Table_sample=Table_sample,
    CN_bin_length=CN_bin_length,
    prob_CN_whole_genome_duplication=prob_CN_whole_genome_duplication,
    prob_CN_missegregation=prob_CN_missegregation,
    prob_CN_chrom_arm_missegregation=prob_CN_chrom_arm_missegregation,
    prob_CN_focal_amplification=prob_CN_focal_amplification,
    prob_CN_focal_deletion=prob_CN_focal_deletion,
    prob_CN_cnloh_interstitial=prob_CN_cnloh_interstitial,
    prob_CN_cnloh_terminal=prob_CN_cnloh_terminal,
    prob_CN_focal_amplification_length=prob_CN_focal_amplification_length,
    prob_CN_focal_deletion_length=prob_CN_focal_deletion_length,
    prob_CN_cnloh_interstitial_length=prob_CN_cnloh_interstitial_length,
    prob_CN_cnloh_terminal_length=prob_CN_cnloh_terminal_length,
    rate_driver=rate_driver,
    rate_passenger=rate_passenger,
    selection_model=selection_model,
    bound_driver=bound_driver,
    bound_average_ploidy=bound_average_ploidy,
    bound_homozygosity=bound_homozygosity,
    table_population_dynamics=table_population_dynamics,
    gc=gc,
    gc_slope=gc_slope,
    gc_int=gc_int,
    sigma1=sigma1,
    num_reads=num_reads
)
vec_driver_genes  <- c('TP53','BRCA1','BRCA2','MSH2','MSH6','ARID1A',
    'STK11','BRAF','MLH1','PIK3R1','ERBB2','CTNNB1','AKT1','PPP2R1A')
vec_driver_role   <- c('TSG','TSG','TSG','TSG','TSG','TSG',
    'TSG','ONCOGENE','TSG','TSG','ONCOGENE','ONCOGENE','ONCOGENE','TSG')
########################################################################
########################################################################
########################################################################
vec_driver_s      <- rep(0.005,length(vec_driver_genes))
# vec_driver_s[1]   <- 0.009


# vec_driver_s      <- rep(0.007,length(vec_driver_genes))
# vec_driver_s[1]   <- 0.01
# vec_driver_s[2]   <- 0.009
# vec_driver_s[3]   <- 0.009
########################################################################
########################################################################
########################################################################
model_variables   <- BUILD_driver_library(model_variables=model_variables,
    vec_driver_genes=vec_driver_genes,
    vec_driver_role=vec_driver_role,
    vec_driver_s=vec_driver_s)
cell_count        <- 20
CN_matrix         <- BUILD_cn_normal_XX(model_variables$cn_info)
drivers           <- list()
model_variables   <- BUILD_initial_population(model_variables=model_variables,
    cell_count=cell_count,
    CN_matrix=CN_matrix,
    drivers=drivers)
model_name        <- 'HMMCOPY-STUDY'
SAVE_model_variables(model_name=model_name,model_variables=model_variables)
```

Create one simulation:

```{r}
start_time = Sys.time()
n_simulations     <- 1
stage_final       <- 3
n_clones_min      <- 1
n_clones_max      <- Inf
simulator_full_program(
    model = model_name,
    n_simulations = n_simulations,
    stage_final = stage_final,
    n_clones_min = n_clones_min,
    n_clones_max = n_clones_max,
    save_simulation = TRUE,
    internal_nodes_cn_info = FALSE,
    save_newick_tree = FALSE,
    save_cn_profile = TRUE,
    model_readcount = FALSE,
    format_cn_profile = "both",
    apply_HMM = FALSE,
    apply_UMAP_on_HMM = FALSE,
    report_progress = TRUE,
    seed = 10,
    compute_parallel = FALSE
)
end_time = Sys.time()
print(end_time - start_time)
```

Make plots for this "original" simulation:

```{r}
plot_all(model = model_name,n_simulations = n_simulations)
```

# Create noisy readcounts with varying parameters

Rename the "original" simulation and its plots:

```{r}
file.rename(paste(model_name, "_simulation_1.rda", sep = ""), paste(model_name, "_simulation_original.rda", sep = ""))
file.rename(paste(model_name, "_sim1_clonal_fishplot.jpeg", sep = ""), paste(model_name, "_original_clonal_fishplot.jpeg", sep = ""))
file.rename(paste(model_name, "_sim1_clonal_phylo.jpeg", sep = ""), paste(model_name, "_original_clonal_phylo.jpeg", sep = ""))
file.rename(paste(model_name, "_sim1_CN_minor_cnTRUTH_phyloTRUTH.jpeg", sep = ""), paste(model_name, "_original_CN_minor.jpeg", sep = ""))
file.rename(paste(model_name, "_sim1_CN_total_cnTRUTH_phyloTRUTH.jpeg", sep = ""), paste(model_name, "_original_CN_total.jpeg", sep = ""))
file.rename(paste(model_name, "_sim1_tot_pop.jpeg", sep = ""), paste(model_name, "_original_tot_pop.jpeg", sep = ""))
```

Load the "original" simulation:

```{r}
load(paste(model_name, "_simulation_original.rda", sep = ""))
simulation_original <- simulation
```

Create a folder as workplace to store files containing noisy readcounts and HMMcopy results:

```{r}
dir.create(model_name)
```

Save data for GC content and mappability as wig files to be used during HMMcopy:

```{r}
SIMULATOR_VARIABLES_for_simulation(model_name)
p2_write_gc_map_as_wig(filename_gc = paste(model_name, "_gc.wig", sep = ""), filename_map = paste(model_name, "_map.wig", sep = ""))
```

Set up variables for the noisy readcount model:

```{r}
table_gc            <<- read.table(paste(model_name, "-input-gc.csv", sep = ""), header = TRUE, sep = ",")
gc_int              <<- 0
gc_slope            <<- 1
vec.sigma1          <- c(seq(1,9,by=2)*1e-2, rep(0.02642392,length=5))
vec.num_reads       <- c(rep(3906632,length=5), seq(1,9,by=2)*1e6)
n_trials            <- length(vec.sigma1)
```

Create noisy readcounts for each parameter set:

```{r}
ind                         <- 0
for (ind in 1:length(vec.sigma1)){
    sigma1                  <<- vec.sigma1[ind]
    num_reads               <<- vec.num_reads[ind]
    #   Create noisy readcounts
    simulation              <- p2_readcount_model(simulation_original, TRUE)
    simulation              <- p2_readcount_model_wide(simulation, TRUE)
    #   Save noisy readcounts
    filename_ind            <- paste(model_name, "_simulation_", ind,".rda", sep = "")
    save(simulation, file = filename_ind)
    sample_cell_ID          <- simulation$sample$sample_cell_ID
    noisy_cn_profiles_long  <- simulation$sample$noisy_cn_profiles_long
    noisy_cn_profiles_wide  <- simulation$sample$noisy_cn_profiles_wide
    filename                <- paste(model_name, "_noisy_cn_profiles_long_", ind, ".csv", sep = "")
    write.csv(noisy_cn_profiles_long, filename, row.names = FALSE)
    filename                <- paste(model_name, "_noisy_cn_profiles_wide_", ind, ".csv", sep = "")
    write.csv(noisy_cn_profiles_wide, filename, row.names = FALSE)
    for (cell in 1:length(sample_cell_ID)) {
        cell_ID             <- sample_cell_ID[cell]
        filename            <- paste(model_name, "/", model_name, "_noisy_cn_profiles_long_", ind, "_", cell_ID, ".wig", sep = "")
        p2_write_cn_as_wig(filename, noisy_cn_profiles_long, cell_ID)
    }
}
```

# Run HMMcopy on noisy readcounts

Run file `hmmcopy.bash` which runs HMMcopy for each parameter set:

```{r}
#===========>   RUN ON GINSBURG:
#   Step 1:     Copy hmmcopy_v0.0.45.sif to folder HMMCOPY-STUDY (or whatever workplace)
#   Step 2:     Change model_name & n_simulations in ginsburg_hmmcopy.sh
#   Step 2:     Copy folder vignettes to Ginsburg/Terremoto
#   Step 3:     sbatch ginsburg_hmmcopy.sh
#===========>   HOW hmmcopy_v0.0.45.sif WAS CREATED:
#   On Ginsburg/Terremoto:
#               module load singularity
#               singularity pull docker://quay.io/mondrianscwgs/hmmcopy:v0.0.45





#===========    IF RUNNING ON COMPUTER: (EXTREMELY SLOW ON MAC M1)
# start_time = Sys.time()
# n_trials            <- length(vec.gc_slope)*length(vec.sigma1)
# flag_parallel       <- 1
# system("chmod -x hmmcopy.bash")
# system(paste("bash hmmcopy.bash ", model_name, " ", n_trials, " ", flag_parallel, sep = ""))
# end_time = Sys.time()
# print(end_time - start_time)
```

Incorporate HMMcopy results back into simulation files:

```{r}
append_with_hmm(model = model_name, n_simulations = n_trials, UMAP = TRUE, pseudo_corrected_readcount = FALSE)
```

Plot the GC-readcount plots:

```{r}
plot_gc_readcount(
    model = model_name,
    n_simulations = n_trials,
    width = 1000,
    height = 1000
)
```

Plot the total CN profiles inferred from HMMcopy:

```{r}
plot_cn_heatmap(
    model = model_name,
    n_simulations = n_trials,
    plotcol = "total-copy",
    CN_data = "HMM",
    phylo = "TRUTH",
    width = 1000,
    height = 1000
)

plot_cn_heatmap(
    model = model_name,
    n_simulations = n_trials,
    plotcol = "total-copy",
    CN_data = "HMM",
    phylo = "UMAP",
    width = 1000,
    height = 1000
)
```

Change filenames into something more recognizable:

```{r}
for (ind in 1:length(vec.sigma1)){
    sigma1                  <- vec.sigma1[ind]
    num_reads               <- vec.num_reads[ind]
    filename_old            <- paste(model_name,'_sim',ind,'_reads_vs_GC.jpeg',sep='')
    filename_new            <- paste('sig=',sigma1,'_cov=',num_reads,'_reads_vs_GC.jpeg',sep='')
    file.rename(filename_old, filename_new)
    filename_old            <- paste(model_name,'_sim',ind,'_CN_total_cnHMM_phyloTRUTH.jpeg',sep='')
    filename_new            <- paste('sig=',sigma1,'_cov=',num_reads,'_cn=HMM_phylo=TRUTH.jpeg',sep='')
    file.rename(filename_old, filename_new)
    filename_old            <- paste(model_name,'_sim',ind,'_CN_total_cnHMM_phyloUMAP.jpeg',sep='')
    filename_new            <- paste('sig=',sigma1,'_cov=',num_reads,'_cn=HMM_phylo=UMAP.jpeg',sep='')
    file.rename(filename_old, filename_new)
}
```

Find statistics for inferred CN from HMMcopy compared to true CN:

```{r}
hmm_stats   <- statistics_hmm(model = model_name, n_simulations = n_trials)
```

Plot the distribution of cells with right & wrong inferred ploidy from HMMcopy:

```{r}
filename <- paste(model_name, "_inferred_ploidy_in_HMM_var=sigma.jpeg", sep = "")
jpeg(file = filename, width = 2000, height = 1000)
df <- data.frame(
    Sigma = rep(vec.sigma1[1:5], 2),
    Percentage = c(
        100 * hmm_stats$ploidy_small[1:5] / (hmm_stats$ploidy_small[1:5] + hmm_stats$ploidy_right[1:5] + hmm_stats$ploidy_big[1:5]),
        100 * hmm_stats$ploidy_big[1:5] / (hmm_stats$ploidy_small[1:5] + hmm_stats$ploidy_right[1:5] + hmm_stats$ploidy_big[1:5])
    ),
    Group = c(rep("Inferred ploidy is too small", 5), rep("Inferred ploidy is too big", 5))
)
df$Sigma <- factor(df$Sigma, levels = vec.sigma1[1:5])
p <- ggplot(
    data = df,
    aes(x = Sigma, y = Percentage, fill = Group)
) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
    theme(text = element_text(size = 40), legend.position = "bottom")
print(p)
dev.off()

filename <- paste(model_name, "_inferred_cn_in_HMM_var=sigma.jpeg", sep = "")
jpeg(file = filename, width = 2000, height = 1000)
df <- data.frame(
    Sigma = rep(vec.sigma1[1:5], 2),
    Percentage = c(
        100 * hmm_stats$cn_small[1:5] / (hmm_stats$cn_small[1:5] + hmm_stats$cn_right[1:5] + hmm_stats$cn_big[1:5]),
        100 * hmm_stats$cn_big[1:5] / (hmm_stats$cn_small[1:5] + hmm_stats$cn_right[1:5] + hmm_stats$cn_big[1:5])
    ),
    Group = c(rep("Inferred CN is too small", 5), rep("Inferred CN is too big", 5))
)
df$Sigma <- factor(df$Sigma, levels = vec.sigma1[1:5])
p <- ggplot(
    data = df,
    aes(x = Sigma, y = Percentage, fill = Group)
) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
    theme(text = element_text(size = 40), legend.position = "bottom")
print(p)
dev.off()

filename <- paste(model_name, "_inferred_ploidy_in_HMM_var=coverage.jpeg", sep = "")
jpeg(file = filename, width = 2000, height = 1000)
df <- data.frame(
    Coverage = rep(vec.num_reads[6:10], 2),
    Percentage = c(
        100 * hmm_stats$ploidy_small[6:10] / (hmm_stats$ploidy_small[6:10] + hmm_stats$ploidy_right[6:10] + hmm_stats$ploidy_big[6:10]),
        100 * hmm_stats$ploidy_big[6:10] / (hmm_stats$ploidy_small[6:10] + hmm_stats$ploidy_right[6:10] + hmm_stats$ploidy_big[6:10])
    ),
    Group = c(rep("Inferred ploidy is too small", 5), rep("Inferred ploidy is too big", 5))
)
df$Coverage <- factor(df$Coverage, levels = vec.num_reads[6:10])
p <- ggplot(
    data = df,
    aes(x = Coverage, y = Percentage, fill = Group)
) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
    theme(text = element_text(size = 40), legend.position = "bottom")
print(p)
dev.off()

filename <- paste(model_name, "_inferred_cn_in_HMM_var=coverage.jpeg", sep = "")
jpeg(file = filename, width = 2000, height = 1000)
df <- data.frame(
    Coverage = rep(vec.num_reads[6:10], 2),
    Percentage = c(
        100 * hmm_stats$cn_small[6:10] / (hmm_stats$cn_small[6:10] + hmm_stats$cn_right[6:10] + hmm_stats$cn_big[6:10]),
        100 * hmm_stats$cn_big[6:10] / (hmm_stats$cn_small[6:10] + hmm_stats$cn_right[6:10] + hmm_stats$cn_big[6:10])
    ),
    Group = c(rep("Inferred CN is too small", 5), rep("Inferred CN is too big", 5))
)
df$Coverage <- factor(df$Coverage, levels = vec.num_reads[6:10])
p <- ggplot(
    data = df,
    aes(x = Coverage, y = Percentage, fill = Group)
) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
    theme(text = element_text(size = 40), legend.position = "bottom")
print(p)
dev.off()
```
