```{r setup}
library(readxl)
library(mosaic)
library(ggplot2)
#-----------------------------------------------------------------------
setwd("/Users/dinhngockhanh/CancerSimulator/R/")
files_sources <- list.files(pattern = "*.r$")
sapply(files_sources, source)
#-----------------------------------------------------------------------
# devtools::install_github("dinhngockhanh/CancerSimulator")
# library(CancerSimulator)
```

First, we set up the fixed parameters for the simulator.

```{r}
cell_lifespan <- 1
T_0 <- list(0, "day")
T_end <- list(300, "day")
Table_sample <- data.frame(Sample_ID = c("SA"), Cell_count = c(1000), Age_sample = c(T_end[[1]]))

T_tau_step <- cell_lifespan / 2
CN_bin_length <- 500000

selection_model <- "chrom-arm-selection"

prob_CN_whole_genome_duplication <- 1e-3
prob_CN_missegregation <- 1e-2
prob_CN_chrom_arm_missegregation <- 0
prob_CN_focal_amplification <- 0
prob_CN_focal_deletion <- 0
prob_CN_cnloh_interstitial <- 0
prob_CN_cnloh_terminal <- 0
model_CN_focal_amplification_length <- "beta"
model_CN_focal_deletion_length <- "beta"
prob_CN_focal_amplification_length_shape_1 <- 0.758304780825031
prob_CN_focal_amplification_length_shape_2 <- 5.33873409782625
prob_CN_focal_deletion_length_shape_1 <- 0.814054548726361
prob_CN_focal_deletion_length_shape_2 <- 6.16614890284825
prob_CN_cnloh_interstitial_length <- 0.005
prob_CN_cnloh_terminal_length <- 0.005
rate_driver <- 0
rate_passenger <- 1e-11

bound_driver <- 3
bound_average_ploidy <- 6
bound_maximum_CN <- 8
bound_homozygosity <- 0

vec_time <- T_0[[1]]:T_end[[1]]
vec_cell_count <- rep(1000,length(vec_time))
table_population_dynamics <- cbind(vec_time, vec_cell_count)

gc <- read.csv(file = system.file("extdata", "gc_map_500kb.csv", package = "CancerSimulator"))
gc_slope <- 1.2
gc_int <- 0
sigma1 <- 0.02642392
num_reads <- 3906632

model_variables_base <- BUILD_general_variables(
    cell_lifespan = cell_lifespan,
    T_0 = T_0, T_end = T_end, T_tau_step = T_tau_step,
    Table_sample = Table_sample,
    CN_bin_length = CN_bin_length,

    prob_CN_whole_genome_duplication = prob_CN_whole_genome_duplication,
    prob_CN_missegregation = prob_CN_missegregation,
    prob_CN_chrom_arm_missegregation = prob_CN_chrom_arm_missegregation,
    prob_CN_focal_amplification = prob_CN_focal_amplification,
    prob_CN_focal_deletion = prob_CN_focal_deletion,
    prob_CN_cnloh_interstitial = prob_CN_cnloh_interstitial,
    prob_CN_cnloh_terminal = prob_CN_cnloh_terminal,

    model_CN_focal_amplification_length = model_CN_focal_amplification_length,
    model_CN_focal_deletion_length = model_CN_focal_deletion_length,
    prob_CN_focal_amplification_length_shape_1 = prob_CN_focal_amplification_length_shape_1,
    prob_CN_focal_amplification_length_shape_2 = prob_CN_focal_amplification_length_shape_2,
    prob_CN_focal_deletion_length_shape_1 = prob_CN_focal_deletion_length_shape_1,
    prob_CN_focal_deletion_length_shape_2 = prob_CN_focal_deletion_length_shape_2,
    prob_CN_cnloh_interstitial_length = prob_CN_cnloh_interstitial_length,
    prob_CN_cnloh_terminal_length = prob_CN_cnloh_terminal_length,
    rate_driver = rate_driver,
    rate_passenger = rate_passenger,
    selection_model = selection_model,
    bound_driver = bound_driver,
    bound_average_ploidy = bound_average_ploidy,
    bound_maximum_CN = bound_maximum_CN,
    bound_homozygosity = bound_homozygosity,
    table_population_dynamics = table_population_dynamics,
    gc = gc,
    gc_slope = gc_slope,
    gc_int = gc_int,
    sigma1 = sigma1,
    num_reads = num_reads
)

arm_id <- c(paste(model_variables_base$cn_info$Chromosome, "p", sep = ""), paste(model_variables_base$cn_info$Chromosome, "q", sep = ""))
arm_chromosome <- rep(model_variables_base$cn_info$Chromosome, 2)
arm_start <- c(rep(1, length(model_variables_base$cn_info$Chromosome)), model_variables_base$cn_info$Centromere_location + 1)
arm_end <- c(model_variables_base$cn_info$Centromere_location, model_variables_base$cn_info$Bin_count)
arm_s <- rep(1, length(arm_id))
arm_s[which(arm_id == "1p")] <- 0.886307781235821
arm_s[which(arm_id == "2p")] <- 1.01825968094413
arm_s[which(arm_id == "3p")] <- 0.920530554894643
arm_s[which(arm_id == "4p")] <- 0.976827677204689
arm_s[which(arm_id == "5p")] <- 1.15969179661142
arm_s[which(arm_id == "6p")] <- 1.04997702081472
arm_s[which(arm_id == "7p")] <- 1.13777132552215
arm_s[which(arm_id == "8p")] <- 0.884012569028591
arm_s[which(arm_id == "9p")] <- 0.935174629285682
arm_s[which(arm_id == "10p")] <- 1.02799999379291
arm_s[which(arm_id == "11p")] <- 0.967134549774896
arm_s[which(arm_id == "12p")] <- 1.06959509411319
arm_s[which(arm_id == "13p")] <- 1
arm_s[which(arm_id == "14p")] <- 1
arm_s[which(arm_id == "15p")] <- 1
arm_s[which(arm_id == "16p")] <- 1.04513229635198
arm_s[which(arm_id == "17p")] <- 0.864520777388622
arm_s[which(arm_id == "18p")] <- 1.01592152467124
arm_s[which(arm_id == "19p")] <- 1.02814969243796
arm_s[which(arm_id == "20p")] <- 1.00368324400257
arm_s[which(arm_id == "21p")] <- 1
arm_s[which(arm_id == "22p")] <- 1
arm_s[which(arm_id == "Xp")] <- 1
arm_s[which(arm_id == "Yp")] <- 1

arm_s[which(arm_id == "1q")] <- 1.18666462081686
arm_s[which(arm_id == "2q")] <- 0.962046964747096
arm_s[which(arm_id == "3q")] <- 1.11600788103894
arm_s[which(arm_id == "4q")] <- 0.937701536089568
arm_s[which(arm_id == "5q")] <- 0.896394926675205
arm_s[which(arm_id == "6q")] <- 0.886499991701187
arm_s[which(arm_id == "7q")] <- 1.05243171270735
arm_s[which(arm_id == "8q")] <- 1.22076350654396
arm_s[which(arm_id == "9q")] <- 0.964639963793044
arm_s[which(arm_id == "10q")] <- 0.90108395332905
arm_s[which(arm_id == "11q")] <- 0.967049029726524
arm_s[which(arm_id == "12q")] <- 1.03537911097717
arm_s[which(arm_id == "13p")] <- 0.969546453082048
arm_s[which(arm_id == "14p")] <- 0.971985635416426
arm_s[which(arm_id == "15p")] <- 0.998823078653965
arm_s[which(arm_id == "16q")] <- 0.935335591311341
arm_s[which(arm_id == "17q")] <- 1.05238889818635
arm_s[which(arm_id == "18q")] <- 0.935292135865439
arm_s[which(arm_id == "19q")] <- 1.03037910158697
arm_s[which(arm_id == "20q")] <- 1.20848277950342
arm_s[which(arm_id == "21q")] <- 0.96470320220417
arm_s[which(arm_id == "22q")] <- 0.959601745948745
arm_s[which(arm_id == "Xq")] <- 1
arm_s[which(arm_id == "Yq")] <- 1

model_variables_base <- BUILD_driver_library(
    model_variables = model_variables_base,
    table_arm_selection_rates = data.frame(Arm_ID = arm_id, Chromosome = arm_chromosome, Bin_start = arm_start, Bin_end = arm_end, s_rate = arm_s)
)

cell_count <- 1000
CN_matrix <- BUILD_cn_normal_XX(model_variables_base$cn_info)
drivers <- list()
model_variables_base <- BUILD_initial_population(
    model_variables = model_variables_base,
    cell_count = cell_count,
    CN_matrix = CN_matrix,
    drivers = drivers
)
model_prefix <- "GROWTHMODE-vs-SIZE"
folder_workplace <- "GROWTHMODE-vs-SIZE"
```

We set up parameter sweeps for the growth mode, ranging from constant (growth mode 1) to exponential with high growth rate, and average cell population size.
For every parameter combination, we create 1000 simulations.
We set possible values for the average cell population size to be from 200, 400,...1800, 2000.

```{r}
n_simulations_per_batch <- 1000

var1_name <- "vec_cell_count"
P0_vals <- seq(10, 1, by = -1) / 10
MASS <- T_end[[1]]
fr <- function(growthrate) {
    if (growthrate != 0) {
        mass <- P0_val / growthrate * (exp(growthrate * T_end[[1]]) - 1)
    } else {
        mass <- P0_val * T_end[[1]]
    }
    return(mass - MASS)
}
for (i in 1:length(P0_vals)) {
    P0_val <- P0_vals[i]
    growthrate <- findZeros(fr(g) ~ g, xlim = c(-1, 1))[[1]][1]
    vec_cell_count <- P0_val * exp(growthrate * vec_time)
    if (i == 1) {
        var1_vals <- vec_cell_count
    } else {
        var1_vals <- cbind(var1_vals, vec_cell_count)
    }
}
var1_labs <- paste0(1:length(P0_vals))

var2_name <- "scale_cell_count"
var2_vals <- seq(200, 2000, by = 200)
```

Here we can make a graph of the total cell count vs. time for each growth mode.
Growth mode 1 is constant.
Growth mode 10 is exponential with high growth rate

```{r}
for (i in 1:length(P0_vals)) {
    df <- data.frame(t = vec_time, value = var1_vals[, i], mode = var1_labs[i])
    if (i == 1) {
        df_all <- df
    } else {
        df_all <- rbind(df_all, df)
    }
}
df_all$mode <- factor(df_all$mode, levels = var1_labs)
filename <- paste0(model_prefix, "_0_growth_mode.jpeg")
jpeg(file = filename, width = 1000, height = 1100)
p <- ggplot(data = df_all, aes(x = t, y = value)) +
    geom_line(aes(colour = mode), size = 2) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
    xlab("Time") +
    ylab("Total cell count") +
    labs(color = "Growth mode") +
    theme(panel.background = element_rect(fill = "white", colour = "grey50"), text = element_text(size = 40), legend.position = "top", legend.justification = "right", legend.direction = "horizontal", legend.key.width = unit(2.5, "cm"), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    guides(colour = guide_legend(nrow = 2))
print(p)
dev.off()
```

We make the simulation run through the possible combinations between the two variables with each single variable combination consisting of 1000 simulations.

```{r}
simulator_multivar(
    model_prefix = model_prefix, folder_workplace = folder_workplace,
    model_variables_base = model_variables_base,
    var1_name = var1_name, var1_vals = var1_vals, var1_labs = var1_labs,
    var2_name = var2_name, var2_vals = var2_vals,
    n_simulations = n_simulations_per_batch,
    stage_final = 3,
    compute_parallel = TRUE
)
```

We obtain the average statistics from the simulations for each parameter value combination.

```{r}
statistics_multivar_matrix(
    model_prefix = model_prefix, folder_workplace = folder_workplace,
    var1_name = var1_name, var1_vals = var1_vals, var1_labs = var1_labs,
    var2_name = var2_name, var2_vals = var2_vals,
    n_simulations = n_simulations_per_batch,
    plot_WGD = TRUE, plot_misseg = TRUE,
    compute_parallel = TRUE
)
```

Here we can see how the statistics change by keeping the average cell population size constant for all different growth modes.

```{r}
var2_fixed <- var2_vals[length(var2_vals)]
statistics_multivar_vector(
    model_prefix = model_prefix, folder_workplace = folder_workplace,
    var1_name = var1_name, var1_labs = var1_labs,
    var2_name = var2_name, var2_vals = rep(var2_fixed, length(var1_vals)),
    var_labs = var1_labs, name_lab = "Growth mode",
    plot_WGD = TRUE,
    n_simulations = n_simulations_per_batch,
    plotname = paste0(model_prefix, "_growth_mode_varies_and_average_cell_count=", var2_fixed),
    example_simulation = FALSE,
    R_libPaths = "/burg/iicd/users/knd2127/rpackages"
)
```

Here we can see how the statistics change by keeping the growth mode constant for all average cell population sizes.

```{r}
var1_fixed <- var1_labs[length(var1_labs)]
statistics_multivar_vector(
    model_prefix = model_prefix, folder_workplace = folder_workplace,
    var1_name = var1_name, var1_labs = rep(var1_fixed, length(var2_vals)),
    var2_name = var2_name, var2_vals = var2_vals,
    var_labs = scientific(var2_vals), name_lab = "Average cell count",
    plot_WGD = TRUE,
    n_simulations = n_simulations_per_batch,
    plotname = paste0(model_prefix, "_average_cell_count_varies_and_growth_mode=", var1_fixed),
    example_simulation = FALSE
)
```
